What We Eval in the Shadows

### 2021-08-13 paper/main.tex
Fix eval usage

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. In this section we show 9 real-world examples from this batch. The
goal is to illustrate the variety of use cases that employ \eval in the R
ecosystem. The code snippets are simplified for clarity.

@@

\paragraph{Code Generation}
\Eval can be used to generate boiler plate code that would either be tedious to
write by hand or that is only known after deployment. The example uses \eval to
automatically generate R bindings to C++ methods.

@@

\paragraph{Extracting Varargs}
\Eval can be used to obtain the source code of the arguments passed to a vararg
parameter. One example is a function that iterates over the contents of the
vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
expression that represents a call to \k{alist} but where the \k{...} are
expended with the arguments supplied to the function. \Eval then returns a
\k{list} with unevaluated argument expressions which are evaluated sequentially
for their result. The reason \eval is needed is not to force
any more arguments once a non-null value was found.

\begin{minipage}{.95\textwidth}
\medskip\underline{\c{statnet.common::NVL}:}
\begin{lstlisting}
 function(...) {
   for(e in eval(substitute(alist(...)))) {
     x <- eval(e, parent.frame())
     if(!is.null(x)) break
   }
   x
 }
\end{lstlisting}\medskip
\end{minipage}

@@

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is
expended via \eval and \k{match.call}. The example uses \k{match.call} to
access the expression with which the function is called. The next three lines
construct a call to \k{stats::model.frame} with argument text of parameters
\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression is
evaluated in the parent environment. The use of \k{match.call}, avoids repeated
use of \k{substitute} for extracting the argument expressions. It is extensible
as new parameter names can be easily added. This example illustrates one of the
most recurring usage of \eval that is used by the majority of packages that do
statistical modeling, \ie the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

@@

The most probable reason for this is to ease maintenance of such
wrappers.\footnote{This was speculated by the Brian Ripley, the author of the
function, in a private conversation.} Wrapping functions in R is somewhat
harder than in other languages because of the default argument values, the
varargs and the missing arguments.

>>>>

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. We present examples from nine classes from this batch. The
classification is not exhaustive; we only study 117 call sites from the
\PkgEvalCallSites call sites in our corpus and many uses of \eval do not fit in
a specific category. The goal here is to demonstrate the wide variety of uses to
which \eval is put to use. This also speaks to the ubiquity of \eval in the R
ecosystem. The code snippets have been simplified for clarity.

@@

\paragraph{Code Generation}
\Eval can be used to generate boilerplate code that would either be tedious to
write by hand or that is only known after deployment. The example uses \eval to
automatically generate R bindings to C++ methods.

@@

\paragraph{Extracting Varargs}
\Eval can be used to obtain the source code of the arguments passed to a vararg
parameter. One example is a function that iterates over the contents of the
vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
expression that represents a call to \k{alist} but where the \k{...} are
expended with the arguments supplied to the function. \Eval then returns a
\k{list} with the unevaluated argument expressions.

\begin{minipage}{.95\textwidth}
\medskip\underline{\c{statnet.common::NVL}:}
\begin{lstlisting}
 function(...) {
   exprs <- eval(substitute(alist(...)))
\end{lstlisting}\medskip
\end{minipage}

@@

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is expended
via \eval and \k{match.call}. The example uses \k{match.call} to access the
expression with which the function is called. The next three lines construct a
call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
parent environment. The use of \k{match.call} avoids repeated use of
\k{substitute} for extracting the argument expressions. It is extensible as new
parameter names can be easily added. This example illustrates one of the most
recurring uses of \eval by packages that do statistical modeling, \ie the
\c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

@@

The most probable reason for this is to ease maintenance of such
wrappers.\footnote{Private communication with a developer of R.} Wrapping
functions in R is somewhat harder than in other languages because of default
arguments, varargs and missing arguments.
